The Design, Semantics, and Implementation of the Synchronous Language \CEU

### 2015-10-10 paper.tex
Typos
Revisao geral

% Title portion
\title{The Design, Semantics, and Implementation of the Synchronous Reactive 
Language \CEU}

@@

Despite its success and influence, Esterel has a overly complex semantics that 
requires ingenious static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} 
problems~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible or non-compliant implementations.
%
More importantly, it also affects the programmer's understanding capacity, 
which, after all, have to solve the errors when facing corner cases in 
programs.
%
Another drawback of the Esterel semantics consists of non-deterministic 
execution for intra-reaction statements, which prevents threads to share memory 
and interact with side-effect and stateful system calls.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of \CEU:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers, showing that the 
peculiarities in the semantics of \CEU does not pose practical obstacles.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
Kbytes of SRAM.)}
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, a simple application that blinks three LEDs periodically 
occupies less than 100 bytes and can be completely transmitted in 4 radio 
messages.

@@

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel (a) and \CEU (b) for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 3--10), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit multiple external input signals, which may 
coexist during a reaction.
%
In \CEU, a reaction starts from the occurrence of an external input event and 
programs cannot emit inputs.
Therefore, the occurring input event is unique during the entire reaction, 
resulting in intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln.  5--6) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 5), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 9), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example in Figure~\ref{lst.await} illustrates this distinction, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then periodically, from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

@@

Embedded applications make extensive use of global memory and shared resources, 
such as through memory-mapped registers and system calls to drivers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory and interacting with the 
environment through system calls.

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
For this reason, Esterel assumes that external calls to the host language do 
not perform side effects.
%
Therefore, Esterel syntactically forbids sharing memory between lines of 
execution to preserve determinism: \emph{``if a variable is written by some 
thread, then it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the example of Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
(ln.  5,8) can never run concurrently, because each trail segment reacts to 
different input events (ln. 4,7), which cannot occur simultaneously (according 
to Section~\ref{sec.ceu.ints}).
%
However, the example of Figure~\ref{lst.shared}.b is non-deterministic, because 
the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
\code{A} (ln. 4,7).

@@

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).%
\footnote{\CEU assumes all system calls perform side effects, unless they are 
annotated as ``\texttt{@pure}''.}
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the example of Figure~\ref{lst.shared}.b is suspicious, 
the assignments to \code{y} are both atomic and deterministic, i.e., after the 
reaction to \code{A} terminates, the value of \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel (a) and \CEU (b) to summarize the semantic difference regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.
%
Note that in Esterel multiple external events can coexist in the same reaction, 
which disallows a similar temporal analysis.

@@

The introductory example of Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution without tweaking them with 
synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
Still, handling abortion when dealing with external resources is challenging 
because they are not subject to the same synchronous execution discipline.

@@

The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
tracks the interaction of \code{par/or} compositions with local variables and 
stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
when the block of \code{buffer} goes out of scope, i.e., the finalization 
clause (after the \code{with}, ln. 18) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

\begin{comment}
The code fragments of Figure~\ref{lst.abortion} shows a corner case regarding
abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
%
For instance, it is not clear in the code in Esterel if the call to \code{f} 
should execute or not after \code{A}, given that the body and abortion events 
are the same.
%
For this reason, Esterel provides \emph{weak} and \emph{strong} variations for 
the \code{abort} statement.
With \emph{strong} abortion (the default), the body is aborted immediately and 
the call does not execute.
%
In \CEU, given the deterministic scheduling rules, strong and weak abortions 
can be chosen by reordering trails inside a \code{par/or}, e.g., in 
Figure~\ref{lst.abortion}, the second trail is strongly aborted by the first 
trail and the call to \code{\_f} never executes.

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example of Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy and takes 15ms to check 
for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the example of Figure~\ref{lst.timers}.b, although the 
scheduler cannot guarantee that the first trail terminates exactly in 11ms, it 
can at least ensure that the program always terminates with \code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the peculiarities of \CEU, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and \C calls (which behave like in conventional 
imperative languages).

@@

The $mem(id)$ primitive represents all accesses, assignments, \C function 
calls, and output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

@@

Most statements from \CEU (``concrete \CEU'') map directly to those presented 
in the reduced syntax of Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
%\code{par/or}.
For instance, the \code{if} in the concrete language behaves exactly like the 
$if$ in the formal semantics.
However, there are some significant mismatches between the concrete and 
abstract \CEU, and we (informally) propose appropriate mappings in this 
section.
%
Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.

@@

\subsection{Memory Layout}
\label{sec:impl:memory}

@@

\subsection{Trail Allocation}
\label{sec:impl:gates}

@@

\subsection{Interaction with the Environment}

>>>>

% Title portion
\title{The Design, Semantics, and Implementation of the Synchronous Language 
\CEU}

@@

Despite its success and influence, Esterel has an overly complex semantics that 
requires ingenious static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems, with extensive coverage in 
the 
literature~\cite{esterel.constructive,esterel.d1,esterel.d2,esterel.d3,esterel.d4,esterel.d5,esterel.d6,esterel.d7,esterel.d8}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Furthermore, it also affects the capacity of the programmer's understanding, 
which, after all, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to share memory and interact with side-effect and stateful system calls 
appropriately.
%
For instance, although Esterel supports orthogonal abortion of 
threads~\cite{esterel.preemption}, it doesn't offer a finalization mechanism to 
release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provide a limited form 
of coroutines.
%
\item A static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item A finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
Kbytes of SRAM.)
}, showing that the peculiarities in the semantics of \CEU does not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, a simple application that blinks three LEDs periodically 
occupies less than 100 bytes and can be transmitted over the air in a couple of 
packets.

@@

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel (a) and \CEU (b) for the following control 
specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit different external input signals 
simultaneously, with all coexisting during a reaction.
%
In \CEU, a reaction starts from an external input event and programs cannot 
emit inputs at all.
Therefore, the occurring input event is unique during the entire reaction, 
resulting in intrinsic queue-based handling.
%
In contrast, programs can emit internal events but these follow a stack-based 
execution policy, similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel (a) and \CEU (b).
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), emits \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

Internal events provide a fine-grained intra-reaction control mechanism.
For instance, it brings a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

\begin{figure}[b]
\begin{lstlisting}[numbers=left,xleftmargin=2em]
event int* inc; // subroutine `inc'
par/or do
    loop do     // definitions are loops
        var int* p = await inc;
        *p = *p + 1;
    end
with
    var int v = 1;
    await A;
    emit inc => &v; // call `inc'
    _assert(v==2);  // after return
end
\end{lstlisting}
\caption{ Subroutine \code{inc} is defined in a loop (ln. 3--6), in parallel 
with the caller (ln. 8--11).
\label{lst.sub}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU has a different take, posing a tradeoff that an \code{await} is only valid 
for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.schizo,esterel.schizo2}).

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
For instance, sometimes we need to execute a block of code immediately, and 
then, periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

@@

Embedded applications make extensive use of global memory and shared resources, 
such as through memory-mapped registers and system calls to drivers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory and interacting with the 
environment.

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
For this reason, Esterel assumes that external calls to the host language do 
not perform side effects.
%
Also, Esterel syntactically forbids sharing memory between lines of execution 
to preserve determinism: \emph{``if a variable is written by some thread, then 
it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the example in Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
(ln.  5,8) can never run concurrently, because each trail segment reacts to 
different input events (ln. 4,7), which cannot occur simultaneously (according 
to Section~\ref{sec.ceu.ints}).
%
However, the example in Figure~\ref{lst.shared}.b is non-deterministic, because 
the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
\code{A} (ln. 4,7).

@@

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).%
\footnote{\CEU assumes that all system calls perform side effects, unless they 
are annotated as ``\texttt{@pure}''.}
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the example in Figure~\ref{lst.shared}.b is suspicious, 
the assignments to \code{y} are both atomic and deterministic, i.e., after the 
reaction to \code{A} terminates, the value of \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares two syntactically equivalent code fragments in 
Esterel (a) and \CEU (b) to summarize the semantic differences regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.
%
Note that in Esterel multiple external events can coexist in the same reaction, 
which disallows a similar temporal analysis.

@@

The introductory example in Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution without tweaking them with 
synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
Still, handling abortion when dealing with external resources is challenging 
because they are not subject to the same synchronous execution discipline.

@@

The unsafe example of Figure~\ref{lst.fin}.a does not compile because \CEU 
tracks the interaction of \code{par/or} compositions with local variables and 
stateful output events calls in order to preserve safe abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
\CEU enforces a \emph{finalization} clause to accompany the output request.
The code in Figure~\ref{lst.fin}.b properly cancels the packet transmission 
when the block of \code{buffer} goes out of scope, i.e., the finalization 
clause (ln. 18) executes automatically on external abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

\begin{comment}
The code fragments in Figure~\ref{lst.abortion} shows a corner case regarding
abortion: when the event \code{A} occurs, the program behavior seems ambiguous.
%
For instance, it is not clear in the code in Esterel if the call to \code{f} 
should execute or not after \code{A}, given that the body and abortion events 
are the same.
%
For this reason, Esterel provides \emph{weak} and \emph{strong} variations for 
the \code{abort} statement.
With \emph{strong} abortion (the default), the body is aborted immediately and 
the call does not execute.
%
In \CEU, given the deterministic scheduling rules, strong and weak abortions 
can be chosen by reordering trails inside a \code{par/or}, e.g., in 
Figure~\ref{lst.abortion}, the second trail is strongly aborted by the first 
trail and the call to \code{\_f} never executes.

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example in Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy and takes 15ms to check 
for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the example in Figure~\ref{lst.timers}.b, although the 
scheduler cannot guarantee that the first trail terminates exactly in 11ms, it 
can at least ensure that the program always terminates with \code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the peculiarities of \CEU, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and system calls (which behave like in conventional 
imperative languages).

@@

The $mem(id)$ primitive represents all accesses, assignments, system calls, and 
output events that affect a memory location identified by $id$.
%
According to the synchronous hypothesis of \CEU, $mem$ expressions are 
considered to be atomic and instantaneous.
%
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%
%The special notation $nop$ is used to represent an innocuous $mem$ expression 
%(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
%unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

@@

Most statements from \CEU (``concrete \CEU'') map directly to those presented 
in the reduced syntax in Figure~\ref{lst.formal.syntax} (``abstract \CEU'').
%, namely, the \code{if}, \code{`;'}, \code{loop}, \code{par/and}, and 
%\code{par/or}.
For instance, the \code{if} in the concrete language behaves exactly like the 
$if$ in the formal semantics.
However, there are some significant mismatches between the concrete and 
abstract \CEU, and we (informally) present appropriate mappings in this 
section.
%
Again, we are not considering side-effects, which are all mapped to the $mem$ 
semantic construct.

@@

\subsection{Static Memory Layout}
\label{sec.impl.data}

@@

\subsection{Static and Lightweight Trail Allocation}
\label{sec.impl.trails}

@@

\subsection{Interaction with the Environment}
\label{sec.impl.environment}
